---
title: "USAID SCMS Health Logistics"
subtitle: "Data Scientist Technical Report"
author: "Data Scientist Track Submission by Murera Gisa"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    code_folding: show
    fig_width: 10
    fig_height: 5.5
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)

# Create directories for output
dir.create("reports/figures", recursive = TRUE, showWarnings = FALSE)
```

## Executive Summary

This report fulfils the **Data Scientist track** of the SFH Rwanda technical challenge.
It covers EDA, data quality and inconsistency checks, data transformation, hypothesis testins and some visual representations from the data.

This report presents an exploratory data analysis (EDA) of the `USAID SCMS Delivery History Dataset`, focusing on __data quality issues__ and testing a key __business hypothesis__ about shipping mode costs. The analysis identifies three critical data quality issues that must be addressed before any modeling/dashboarding or reporting, and provides statistical evidence for air freight being significantly more expensive than truck transport for heavy loads.

Finaly this reports provides the overall reproducible functions to automatically validate the data since it is vital and important for any data-related project.

> Key Findings:

- Issue 1: $\approx 40\%$ of freight cost values are non-numeric text requiring classification

- Issue 2: Extreme outliers ($>\$500,000$) skew freight cost distributions

- Issue 3: "N/A" string values create a spurious shipment mode category

> Hypothesis:

Air freight is $6.5\times$ more expensive than truck for loads $>500$ kg ($p < 0.001$)

## 1. Data Loading and Initial Inspection

```{r}
# Load the raw dataset
raw <- read_csv(here::here("data/raw/SCMS_Delivery_History_Dataset.csv"),
                locale = locale(encoding = "latin1"),  # handles Côte d'Ivoire accent
                show_col_types = FALSE) %>% janitor::clean_names()

# Display basic dataset information
cat(sprintf("  Rows    : %d\n",    nrow(raw)))
cat(sprintf("  Columns : %d\n",    ncol(raw)))
cat(sprintf("  Period  : %s  to  %s\n\n",
            min(raw$scheduled_delivery_date, na.rm = TRUE),
            max(raw$scheduled_delivery_date, na.rm = TRUE)))
```

- *Interpretation:* The dataset contains `r nrow(raw)` shipments spanning from 
`r min(raw$scheduled_delivery_date, na.rm = TRUE)` to
`r max(raw$scheduled_delivery_date, na.rm = TRUE)`. With `r ncol(raw)` variables, this provides a rich foundation for analyzing supply chain efficiency and costs across different shipping modes and destinations.

## 2.Data Quality Issues 

This section presents the data quality issues and present them in interactive way.

### Issue no. 1: Non numeric Freight Cost Values

*State of the Problem:* The freight_cost_usd column contains approximately $40\%$ non-numeric text values. A naive `as.numeric()` conversion would silently convert these to `NA`, losing valuable information and potentially biasing analyses.

```{r}
freight_classes <- tibble(val = raw$`freight_cost_usd`) %>%
  mutate(
    category = case_when(
      str_detect(val, "^\\d")                                ~ "Numeric (parseable)",
      str_detect(val, "Freight Included|Invoiced Separately") ~ "Included in commodity cost",
      str_detect(val, "^See ")                               ~ "Cross-reference to other row",
      is.na(val) | val == ""                                 ~ "Blank / missing",
      TRUE                                                   ~ "Other text"
    )
  ) %>%
  count(category) %>%
  mutate(pct = round(100 * n / nrow(raw), 1)) %>%
  arrange(desc(n))

# Display the breakdown
kable(freight_classes, 
      col.names = c("Category", "Count", "Percentage"),
      caption = "Table 1: Freight Cost Value Categories") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

non_numeric_n <- sum(freight_classes$n[freight_classes$category != "Numeric (parseable)"])

#Visualization 

p1 <- ggplot(freight_classes,
             aes(x = reorder(category, n), y = n, fill = category)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = paste0(n, "  (", pct, "%)")),
            hjust = -0.05, size = 3.6) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Numeric (parseable)"             = "#6DB57A",
    "Included in commodity cost"      = "#F5A623",
    "Cross-reference to other row"    = "#E05C5C",
    "Blank / missing"                 = "#B0B0B0",
    "Other text"                      = "#888888"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.28))) +
  labs(
    title    = "Issue #1: ~40% of Freight Cost Values Are Non-Numeric Text",
    subtitle = "Naive as.numeric() would silently discard these as NA — three categories need different treatment",
    x = NULL, y = "Number of records",
    caption  = "Author: Murera Gisa \n Source: USAID SCMS Delivery History Dataset"
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", colour = "#1A3A5C"))

print(p1)
ggsave("reports/figures/issue1_non_numeric_freight.png", p1, width = 10, height = 4.5, dpi = 150)
non_numeric_n <- sum(freight_classes$n[freight_classes$category != "Numeric (parseable)"])
```

*Interpretation:* The above Figure 1 reveals `r non_numeric_n` records
(`r round(100 * non_numeric_n / nrow(raw), 1)`\%) records
with non-numeric freight costs. These fall into three distinct categories requiring different treatments:

- **Included in commodity cost**: (`r freight_classes$n[3]` == "Included in commodity cost"] records): The freight cost is bundled with product cost, and  these should be set to $\$0$ for separate freight analysis

- **Cross-reference**: to other row" (`r freight_classes$n[2]` == "Cross-reference to other row"] records): Costs are documented in another shipment record ,this requires joining logic

- **Blank or Missing**: The results shows that  `287` records are missing in `line_item_insurance_usd` column. True missing values requiring imputation or exclusion.

**Business Impact:** Ignoring this issue would result in $40\%$ data loss for any freight cost analysis, potentially biasing results toward shipments that explicitly itemize freight costs.

### Issue No.2 Extreme Outliers in Freight Costs

* State of the Problem:*
Among numerically parseable freight costs, the distribution shows extreme right skew with values exceeding $\$500,000$, which would inflate average cost calculations.

```{r}
numeric_freight <- raw %>%
  mutate(freight_usd = suppressWarnings(as.numeric(freight_cost_usd))) %>%
  filter(!is.na(freight_usd), freight_usd >= 0)

q1_val      <- quantile(numeric_freight$freight_usd, 0.25)
q3_val      <- quantile(numeric_freight$freight_usd, 0.75)
iqr_val     <- q3_val - q1_val
upper_fence <- q3_val + 1.5 * iqr_val
outlier_n   <- sum(numeric_freight$freight_usd > upper_fence)

# Create summary statistics table
outlier_stats <- data.frame(
  Metric = c("Numeric Records", "Median Cost", "IQR Upper Fence", 
             "Outliers Above Fence", "Maximum Value"),
  Value = c(nrow(numeric_freight),
            paste0("$", format(round(median(numeric_freight$freight_usd)), big.mark = ",")),
            paste0("$", format(round(upper_fence), big.mark = ",")),
            paste0(outlier_n, " (", round(100 * outlier_n / nrow(numeric_freight), 1), "%)"),
            paste0("$", format(round(max(numeric_freight$freight_usd)), big.mark = ",")))
)

kable(outlier_stats, col.names = c("Metric", "Value"),
      caption = "Table 2: Freight Cost Outlier Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

p2 <- numeric_freight %>%
  filter(freight_usd <= quantile(freight_usd, 0.995)) %>%   # truncate for readability
  ggplot(aes(x = freight_usd)) +
  geom_histogram(bins = 70, fill = "#4A90D9", colour = "white", linewidth = 0.15) +
  geom_vline(xintercept = upper_fence, colour = "#E05C5C",
             linetype = "dashed", linewidth = 1.1) +
  annotate("text", x = upper_fence * 1.06, y = Inf, vjust = 2, hjust = 0,
           label = paste0("IQR fence\n$", format(round(upper_fence), big.mark = ",")),
           colour = "#E05C5C", size = 3.6) +
  scale_x_continuous(labels = dollar_format(big.mark = ",")) +
  labs(
    title    = "Issue #2: Extreme Right Skew — Outliers in Numeric Freight Costs",
    subtitle = "X-axis truncated at 99.5th percentile; long tail reaches >$500k",
    x = "freight_cost_usd", y = "Count",
    caption  = "Author: Murera Gisa \n Source: USAID SCMS Delivery History Dataset"
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", colour = "#1A3A5C"))

print(p2)
ggsave("reports/figures/issue2_freight_outliers.png", p2, width = 10, height = 4.5, dpi = 150)
```

*Interpretation:* The above Figure 2 shows the distribution of numeric freight costs (truncated at the $99.5^{th}$ percentile for visibility). The IQR method identifies 
`r outlier_n` outliers (`r round(100 * outlier_n / nrow(numeric_freight), 1)`\% of numeric records) above the upper fence of `r format(round(upper_fence), big.mark = ",")`. The maximum value of `r format(round(max(numeric_freight$freight_usd)), big.mark = ",")` is orders of magnitude higher than the median.

**Business Impact:** Including these outliers in average cost calculations would inflate estimates by $10$ to $100$ times, potentially leading to:

- Overstated budget requirements

- Incorrect mode cost comparisons

- Misleading country-level cost analyses

**Recommended Treatment:** Flag outliers using IQR method and exclude them from average-cost analyses, or use robust statistics (median, trimmed means).

### Issue no 3. String 'N/A' Sentinel in Shipment Mode

_State of the Problem:_
The literal string `N/A` is used as a sentinel value in the Shipment Mode column. R's `is.na()` returns FALSE for these values, meaning they would be treated as a legitimate fifth shipping mode in analyses.

```{r}
mode_dist <- raw %>%
  count(shipment_mode, sort = TRUE) %>%
  mutate(
    pct         = round(100 * n / nrow(raw), 1),
    is_sentinel = shipment_mode == "N/A"
  )

kable(mode_dist, 
      col.names = c("Shipment Mode", "Count", "Percentage", "Is Sentinel"),
      caption = "Table 3: Shipment Mode Distribution") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

p3 <- ggplot(mode_dist,
             aes(x = reorder(shipment_mode, n), y = n,
                 fill = is_sentinel)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = paste0(n, "  (", pct, "%)")),
            hjust = -0.05, size = 3.8) +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "#1A3A5C", "TRUE" = "#E05C5C")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.22))) +
  labs(
    title    = "Issue #3: 360 Records Have Shipment Mode = 'N/A' (a String, Not R's NA)",
    subtitle = "Red bar = sentinel value; is.na() returns FALSE — must be explicitly recoded",
    x = NULL, y = "Number of records",
    caption  = "Authr: Murera Gisa, \n Source: USAID SCMS Delivery History Dataset"
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", colour = "#1A3A5C"))

print(p3)
ggsave("reports/figures/issue3_shipment_mode_na.png", p3, width = 10, height = 4, dpi = 150)
```

*Interpretation:* Figure 3 highlights `r mode_dist$n[mode_dist$shipment_mode== "N/A"]` records (`r mode_dist$pct[mode_dist$shipment_mode== "N/A"]`\%) where Shipment Mode contains the string "N/A" rather than a true NA value. These would be incorrectly treated as a separate shipping mode in any grouping operation.

**Business Impact:** Without recoding, any analysis by shipping mode would:

- Include a spurious "N/A" category

- Misrepresent the distribution of legitimate shipping modes

- Potentially skew mode-specific cost calculations

**Recommended Treatment:** Explicitly recode "N/A" to NA_character_ before any analysis:

```{r}
raw$Shipment_Mode_clean <- ifelse(raw$shipment_mode == "N/A", NA, raw$shipment_mode)

```

## Hypothesis Test: Air vs Truck Cost-per-KG for Heavy Loads

### Hypothesis Formulation

**Business Question:** For heavy loads which is greater than $500 Kgs$), is air freight significantly more expensive per kilogram than truck transport?

**Statistical Framework:**

- *H₀ (Null):* No significant difference in cost/kg between Air and Truck for load greater than 500 Kgs.

- *H₁ (Alternative):* Air costs is significantly more per kg than Truck for heavy loads

**Test Approach:** To test the above hypotheses we have used *Wilcoxon rank-sum test* which non-parametric test suitable for our skewed cost data.Therefore we have considered the *significance level* of $\alpha = 0.05$ and confidence level of $95\%$

### Data Preparation

```{r}
hyp_df <- raw %>%
  mutate(
    freight_usd = suppressWarnings(as.numeric(freight_cost_usd)),
    weight_kg   = suppressWarnings(as.numeric(weight_kilograms)),
    mode_clean  = case_when(
      shipment_mode == "Air"         ~ "Air",
      shipment_mode == "Air Charter" ~ "Air Charter",
      shipment_mode == "Truck"       ~ "Truck",
      shipment_mode == "Ocean"       ~ "Ocean",
      TRUE                             ~ NA_character_
    )
  ) %>%
  filter(
    !is.na(freight_usd), freight_usd > 0,
    !is.na(weight_kg),   weight_kg > 500,   # heavy loads only
    mode_clean %in% c("Air", "Truck", "Ocean")
  ) %>%
  mutate(cost_per_kg = freight_usd / weight_kg) %>%
  filter(cost_per_kg < quantile(cost_per_kg, 0.99, na.rm = TRUE))  # trim top 1% for visualization

# Descriptive statistics
hyp_summary <- hyp_df %>%
  group_by(Mode = mode_clean) %>%
  summarise(
    n           = n(),
    median_cpk  = round(median(cost_per_kg), 2),
    mean_cpk    = round(mean(cost_per_kg), 2),
    sd_cpk      = round(sd(cost_per_kg), 2),
    .groups     = "drop"
  )

kable(hyp_summary, 
      col.names = c("Mode", "N", "Median ($/kg)", "Mean ($/kg)", "SD ($/kg)"),
      caption = "Table 4: Cost per KG Statistics by Mode (Loads >500 kg)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

### Statistical Analysis

```{r}
# Wilcoxon rank-sum test: one-sided Air > Truck
air_cpk   <- hyp_df$cost_per_kg[hyp_df$mode_clean == "Air"]
truck_cpk <- hyp_df$cost_per_kg[hyp_df$mode_clean == "Truck"]
wt        <- wilcox.test(air_cpk, truck_cpk, alternative = "greater")

# Effect size r = Z / sqrt(N)
z_stat    <- qnorm(wt$p.value / 2, lower.tail = FALSE)
effect_r  <- round(z_stat / sqrt(length(air_cpk) + length(truck_cpk)), 3)
multiplier <- round(median(air_cpk) / median(truck_cpk), 1)

# Create results table
test_results <- data.frame(
  Metric = c("Sample Size (Air)", "Sample Size (Truck)", 
             "Median Air ($/kg)", "Median Truck ($/kg)",
             "Cost Multiplier", "P-value", "Effect Size (r)", 
             "Effect Magnitude", "Decision (α=0.05)"),
  Value = c(length(air_cpk), length(truck_cpk),
            paste0("$", median(air_cpk)), paste0("$", median(truck_cpk)),
            paste0(multiplier, "×"),
            if (wt$p.value < 0.001) "< 0.001" else round(wt$p.value, 4),
            effect_r,
            if (effect_r >= 0.5) "Large" else if (effect_r >= 0.3) "Medium" else "Small",
            if (wt$p.value < 0.05) "REJECT H₀" else "FAIL TO REJECT H₀")
)

kable(test_results, col.names = c("Metric", "Value"),
      caption = "Table 5: Wilcoxon Rank-Sum Test Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

### Visualization

```{r}
p4 <- hyp_df %>%
  filter(mode_clean %in% c("Air", "Truck", "Ocean")) %>%
  ggplot(aes(x = mode_clean, y = cost_per_kg, fill = mode_clean)) +
  geom_violin(alpha = 0.35, colour = NA, trim = TRUE) +
  geom_boxplot(width = 0.22, outlier.alpha = 0.25, colour = "grey30") +
  stat_summary(fun = median, geom = "point",
               shape = 21, size = 4.5, fill = "white", colour = "#1A3A5C") +
  scale_fill_manual(values = c(Air = "#4A90D9", Truck = "#6DB57A", Ocean = "#F5A623"),
                    guide  = "none") +
  scale_y_continuous(labels = dollar_format()) +
  annotate("text", x = 1.5, y = max(hyp_df$cost_per_kg) * 0.93,
           label = paste0("p ", if (wt$p.value < 0.001) "< 0.001" else round(wt$p.value, 3),
                          "\nr = ", effect_r),
           size = 4.2, colour = "#1A3A5C", fontface = "bold") +
  labs(
    title    = paste0("Hypothesis: Air is ", multiplier,
                      "× More Expensive per KG than Truck (Loads > 500 kg)"),
    subtitle = "Wilcoxon rank-sum test: H0 rejected (p < 0.001) | White dot = median",
    x = NULL, y = "Cost per KG (USD)",
    caption  = "Top-1% trimmed for clarity | Author: Murera Gisa \n Source: USAID SCMS Delivery History Dataset"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", colour = "#1A3A5C"))

print(p4)
ggsave("reports/figures/hypothesis_air_vs_truck.png", p4, width = 9, height = 5.5, dpi = 150)

```

### Interpretation

- Statistical Findings:

  - P-value: `r if (wt$p.value < 0.001) "< 0.001" else round(wt$p.value, 4)`: strongly significant

  - Effect Size (r): `r effect_r` — r` if (effect_r >= 0.5) "large" else if (effect_r >= 0.3) "medium" else "small"`  effect

  - Cost Difference: Air freight is `r multiplier` $\times$  more expensive per kg than truck transport (median: `r median(air_cpk)` vs `r median(truck_cpk)`)
  
### Business Interpretation

We have rejected the null hypothesis at $\alpha = 0.05$ with a confidence level of $95\%$. There is strong statistical evidence that air freight costs significantly more per kilogram than truck transport for loads exceeding $500$Kgs. This finding has important implications for supply chain optimization as follow:

- *Cost-Benefit Analysis:* The `r multiplier` $\times$ cost premium for air freight must be justified by time sensitivity or accessibility requirements

- *Mode Selection Criteria:* Organizations should establish clear thresholds where truck transport is preferred unless delivery speed is critical

- *Budget Planning:* Air freight budgets should be `r multiplier`$\times$ higher per kg than truck budgets for comparable heavy shipments

**The Analytical Notes concerning the Ocean Freight:** While included for reference, ocean freight shows the lowest median cost (`r hyp_summary$median_cpk[hyp_summary$Mode == "Ocean"]`/kg), making it the most economical option for non-urgent heavy shipments with adequate lead time.

**CURIOUS QUESTION:** Which is the shipping volume per year?

```{r}
yearly <- raw %>%
  mutate(yr = year(dmy(scheduled_delivery_date))) %>%
  filter(!is.na(yr)) %>%
  count(yr)

kable(yearly, col.names = c("Year", "Shipment Count"),
      caption = "Table 6: Annual Shipment Volumes") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

The table 6 shows that the Peak Year: `r yearly$yr[which.max(yearly$n)]` with 
`r max(yearly$n)` shipments, representing the highest operational volume in the dataset.

## Conclusion

- This EDA has successfully identified three critical data quality issues that must be addressed before proceeding with downstream analysis.

- The hypothesis test confirms, with strong statistical evidence, that air freight carries a significant cost premium over truck transport for heavy loads. The analysis finds that it has direct implications on supply chain optimization.

- The documented issues and recommended solutions provide a clear roadmap for the data cleaning phase, ensuring that subsequent modeling and reporting will be built on a solid, trustworthy foundation.

## Data Story Telling 

This section answers some questions eventhough it is presented in both detailed (`shiny dashbaord`) and superset dashboards

```{r echo=FALSE}
df <- read_csv("data/processed/cleaned_shipments.csv", show_col_types = FALSE)

```

### Question 1: Supply Chain Health

>  Which countries have the highest average freight costs?

```{r echo=FALSE}
q1_data <- df %>%
  filter(freight_category == "Numeric",
         !freight_outlier_flag,
         freight_cost_usd > 0) %>%
  group_by(country) %>%
  summarise(avg_freight = mean(freight_cost_usd, na.rm = TRUE),
            n_shipments = n(), .groups = "drop") %>%
  arrange(desc(avg_freight)) %>%
  slice_head(n = 15)

p_q1 <- ggplot(q1_data,
               aes(x = reorder(country, avg_freight), y = avg_freight)) +
  geom_col(aes(fill = avg_freight), show.legend = FALSE, width = 0.72) +
  geom_text(aes(label = dollar(round(avg_freight))),
            hjust = -0.07, size = 3.5, colour = "grey30") +
  coord_flip() +
  scale_fill_gradient(low = "#7EC8F5", high = "#1A3A5C") +
  scale_y_continuous(labels = dollar_format(),
                     expand = expansion(mult = c(0, 0.22))) +
  labs(
    title    = "Supply Chain Health: Top 15 Countries by Avg Freight Cost",
    subtitle = "Numeric costs only | IQR outliers excluded | 2006–2015",
    x = NULL, y = "Average Freight Cost (USD)",
    caption  = "Author: Murera Gisa \n Source: USAID SCMS Delivery History"
  ) +
  theme_bw()

print(p_q1)
```

### QUESTION 2: Operation Efficiency

> Tuelve months trend of on-time delivey

```{r echo=FALSE}
last_12 <- df %>%
  filter(delivery_status != "Unknown", !is.na(year_month)) %>%
  group_by(year_month) %>%
  summarise(total = n(),
            on_time = sum(delivery_status == "On-Time"),
            ontime_pct = round(100 * on_time / total, 1),
            .groups = "drop") %>%
  arrange(year_month) %>%
  tail(12)

p_q2 <- ggplot(last_12, aes(x = year_month, y = ontime_pct, group = 1)) +
  geom_area(fill = "#4A90D9", alpha = 0.10) +
  geom_line(colour = "#1A3A5C", linewidth = 1.4) +
  geom_point(aes(colour = ontime_pct < 70), size = 4, show.legend = FALSE) +
  scale_colour_manual(values = c("FALSE" = "#6DB57A", "TRUE" = "#E05C5C")) +
  geom_hline(yintercept = 80, linetype = "dashed",
             colour = "grey50", linewidth = 0.9) +
  annotate("text", x = last_12$year_month[1], y = 82,
           label = "80% target", hjust = 0, size = 3.5, colour = "grey60") +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(
    title    = "Operational Efficiency: On-Time Delivery — Last 12 Months",
    subtitle = "Red points = months below 70% | Dashed line = 80% programme target",
    x = NULL, y = "On-Time Delivery Rate (%)",
    caption  = "Author: Murera Gisa \n Source: USAID SCMS Delivery History"
  ) +
  theme_bw()

print(p_q2)
```

### Data Story and Insight

```{r echo=FALSE}
air_cpk   <- median(df$cost_per_kg[df$shipment_mode_clean == "Air"],   na.rm = TRUE)
truck_cpk <- median(df$cost_per_kg[df$shipment_mode_clean == "Truck"], na.rm = TRUE)
mult      <- round(air_cpk / truck_cpk, 1)

air_ot    <- round(100 * mean(df$delivery_status[df$shipment_mode_clean == "Air"]   == "On-Time", na.rm = TRUE), 0)
truck_ot  <- round(100 * mean(df$delivery_status[df$shipment_mode_clean == "Truck"] == "On-Time", na.rm = TRUE), 0)

cat(sprintf("  Air median $/kg   : $%.2f\n", air_cpk))
cat(sprintf("  Truck median $/kg : $%.2f\n", truck_cpk))
cat(sprintf("  Multiplier        : %.1fx\n", mult))
cat(sprintf("  Air on-time       : %d%%\n",  air_ot))
cat(sprintf("  Truck on-time     : %d%%\n\n", truck_ot))

# ── Big Number Chart ───────────────────────────────────────────────────────────
p_bn <- ggplot() +
  annotate("rect", xmin=0, xmax=10, ymin=0, ymax=10, fill="#1A3A5C") +
  annotate("text", x=5, y=7.1, label=paste0(mult, "×"),
           colour="#F5A623", size=32, fontface="bold", hjust=0.5) +
  annotate("text", x=5, y=5.0,
           label="Air freight costs MORE per kg than Truck",
           colour="white", size=7, fontface="bold", hjust=0.5) +
  annotate("text", x=5, y=3.1,
           label=paste0(
             "Air on-time rate (", air_ot, "%)  ≈  Truck (", truck_ot, "%)\n",
             "The cost premium buys NO better delivery performance."
           ),
           colour="#7EC8F5", size=4.8, hjust=0.5, lineheight=1.5) +
  annotate("text", x=5, y=0.7,
           label="USAID SCMS Delivery History | 2006–2015 | 10,324 shipments",
           colour="grey60", size=3.2, hjust=0.5) +
  xlim(0,10) + ylim(0,10) +
  theme_void()

print(p_bn)
```

### Historical Trends

```{r echo=FALSE}
# ── Time-Series Chart ──────────────────────────────────────────────────────────
PALETTE <- c(Air="#4A90D9", Truck="#6DB57A", Ocean="#F5A623", "Air Charter"="#7EC8F5")

ts_cpk <- df %>%
  filter(!is.na(cost_per_kg), !is.na(year),
         !is.na(shipment_mode_clean),
         cost_per_kg < quantile(cost_per_kg, 0.99, na.rm=TRUE)) %>%
  group_by(year, Mode=shipment_mode_clean) %>%
  summarise(avg_cpk=mean(cost_per_kg, na.rm=TRUE), .groups="drop")

ts_ot <- df %>%
  filter(delivery_status != "Unknown", !is.na(year),
         !is.na(shipment_mode_clean)) %>%
  group_by(year, Mode=shipment_mode_clean) %>%
  summarise(ontime_pct=round(100*mean(delivery_status=="On-Time"),1),
            .groups="drop")

p_ts_top <- ggplot(ts_cpk, aes(x=year, y=avg_cpk, colour=Mode, group=Mode)) +
  geom_line(linewidth=1.6) + geom_point(size=3.2) +
  scale_colour_manual(values=PALETTE) +
  scale_y_continuous(labels=dollar_format()) +
  scale_x_continuous(breaks=2006:2015) +
  labs(
    title    = "The Director's Story: Cost per KG by Mode — 10-Year Trend",
    subtitle = paste0(
      "Air consistently costs more per kg across all 10 years. ",
      "For bulk non-urgent commodities, switching to Truck\n",
      "reduces freight spend without sacrificing delivery reliability."
    ),
    x=NULL, y="Avg Cost per KG (USD)", colour="Mode"
  ) +
  theme_bw() + theme(legend.position="top")

p_ts_bot <- ggplot(ts_ot, aes(x=year, y=ontime_pct, colour=Mode, group=Mode)) +
  geom_line(linewidth=1.1, linetype="dashed") + geom_point(size=2.5) +
  scale_colour_manual(values=PALETTE, guide="none") +
  scale_y_continuous(labels=function(x) paste0(x,"%"), limits=c(0,100)) +
  scale_x_continuous(breaks=2006:2015) +
  labs(subtitle="On-time rate by mode — no consistent Air advantage",
       x="Year", y="On-Time Rate (%)") +
  theme_bw()

# Stack the two panels
p_ts_combined <- gridExtra::grid.arrange(p_ts_top, p_ts_bot,
                                         nrow=2, heights=c(3, 1.6))



```

=========================================================================================
## APPENDIX: Full functions for data validation 

```{r}
#' Comprehensive Data Quality Check for SCMS Dataset
#' @param df Raw SCMS dataframe
#' @return List of quality issues and recommendations

check_scms_quality <- function(df) {
  issues <- list()
  
  # Issue 1: Non-numeric freight
  freight_text <- sum(!str_detect(df$`Freight Cost (USD)`, "^\\d") & 
                       !is.na(df$`Freight Cost (USD)`))
  issues$freight_non_numeric <- list(
    count = freight_text,
    pct = round(100 * freight_text / nrow(df), 1),
    recommendation = "Classify into: numeric, 'Included', 'See row', or missing"
  )
  
  # Issue 2: Outliers
  numeric_vals <- suppressWarnings(as.numeric(df$`Freight Cost (USD)`))
  q3 <- quantile(numeric_vals, 0.75, na.rm = TRUE)
  q1 <- quantile(numeric_vals, 0.25, na.rm = TRUE)
  iqr <- q3 - q1
  outliers <- sum(numeric_vals > (q3 + 1.5 * iqr), na.rm = TRUE)
  issues$freight_outliers <- list(
    count = outliers,
    pct = round(100 * outliers / sum(!is.na(numeric_vals)), 1),
    recommendation = "Flag with IQR method; use median for robust averages"
  )
  
  # Issue 3: "N/A" sentinel
  na_strings <- sum(df$`Shipment Mode` == "N/A", na.rm = TRUE)
  issues$mode_na_string <- list(
    count = na_strings,
    pct = round(100 * na_strings / nrow(df), 1),
    recommendation = "Recode 'N/A' to NA_character_ before analysis"
  )
  
  return(issues)
}

```

**Conclusion:** You need only to load the dataset with the same column names as our data [Supply Chain Shipment Pricing Data](https://www.kaggle.com/datasets/apoorvwatsky/supply-chain-shipment-pricing-data/data)

